# Drawing a boxplot

1. You'll learn how to draw basic boxplots, which are useful for comparing groups on a numeric variable.
2. You'll tinker with the appearence of your graphs.
3. You'll learn how to compute basic summaries such as group means.

## Reading in data

1. Download the file `VowelChoices_ij.csv` from
Moodle to the `data` subdirectory of your R project directory. 
This file contains a part of the results
of a learning experiment [@Vanhove2016]
in which 80 participants were randomly assigned to one
of two learning conditions
(`LearningCondition`).
The column `PropCorrect` contains the proportion 
of correct answers for each participant,
and the question we want to answer concerns
the difference in the performance between
learners in the different conditions.
2. Open your R project.
3. Create a new script and copy the commands with which you loaded the packages and imported the dataset from last week's script to this new script.
4. Adapt the command for importing the data so that it'll read in the file `VowelChoices_ij.csv` and that the dataset will be known in R as `d_box`. Execute these commands.

From now on, I'll assume that you won't enter commands directly to the R console but that you'll first enter them into a script which you'll then execute. (**!**)

**Tip:** Comment your code (with `#`). 
This way you'll be able to understand what your code accomplishes in a few months' time, 
which will make it easier to recycle your scripts. 
You don't have to copy my comments verbatim.
_"Your closest collaborator is you six months ago but you don't reply to email."_

```{r, echo = FALSE, warning = FALSE, message = FALSE}
d_box <- read_csv(here("data", "VowelChoices_ij.csv"))
```

5. Inspect the dataset in R to check that it contains 80 rows and 3 labelled columns (`Subject`, `LearningCondition`, `PropCorrect`).

## What are boxplots?
Below you see a basic boxplot of the vocabulary test scores of the
eighty participants in [@Vanhove2016].
The median score is highlighted by a thick line.
The 25th and 75th percentiles 
(also known as the first and third quartiles)
are highlighted by thinner lines and together
form a box. The difference between the 75th and 25th percentile is
called the **inter-quartile range (IQR)**.
The data points that don't fall in the box, that is, the data points that don't
make up the middle 50% of the data, are represented by a line protruding from
the upper part of the box and by a line protruding from the lower part of the box.
However, data points whose distance to the box exceeds 1.5 times the inter-quartile
range are plotted separately. If such data points exist, the lines protruding from 
the box only extend up to the lowest / highest data point whose distance to the 
box is lower than 1.5 times the IQR.

From the boxplot below, we can glean that the IQR is $34 - 31 = 3$.
The highest data point has a value of 38. Since $38 < 34 + 1.5 \times 3 = 38.5$,
this data point is captured by the line protruding from the upper part of the box.
The lowest data point has a value of 25. Since $25 < 31 - 1.5 \times 3 = 26.5$,
it lies too far from the box to be captured by the line protruding from the lower
part of the box, so it gets drawn separately.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dat <- read_csv(here("data", "Vanhove2015_Vpn.csv"))
old_ops <- par()
par(mar = c(1, 4, 2, 1), las = 1, cex = 1.3)
boxplot(dat$Wortschatz, boxwex = 0.25, staplewex = 0,
        ylab = "Vocabulary test score", col = "white",
        main = "Example boxplot", pars = list(boxwex=0.4), at=0.3, xlim=c(0.15,0.55))
text(x = 0.45, max(dat$Wortschatz), "Maximum")
text(x = 0.45, quantile(dat$Wortschatz, .75), "75th percentile", font=3)
text(x = 0.45, median(dat$Wortschatz), "Median", font=2)
text(x = 0.45, median(dat$Wortschatz)-0.8, "(= 50th percentile)")
text(x = 0.45, quantile(dat$Wortschatz, .25), "25th percentile", font=3)
text(x = 0.45, 27, "'Minimum'")
text(x = 0.43, 25.5, "Data points more than\n1.5*IQR from the box")
par(old_ops)
```


## Drawing boxplots
We can draw boxplots in order to compare the distribution of a variable
in two groups.
But you can also use
boxplots to compare more than two groups.
Next week we'll add the individual data points to
these boxplots which can make them more informative still.

From now on, I'll always assume that you've loaded the `tidyverse` and `here` packages.

1. The following command plots the accuracy
data (`PropCorrect`) using a separate boxplot for
each condition (`LearningCondition`). 
You already know how the first three lines work;
the fourth specifies that the data should be plotted
using boxplots.

```{r, fig.asp = 0.5}
# Boxplot PropCorrect vs. LearningCondition
ggplot(data = d_box,              
       aes(x = LearningCondition, 
           y = PropCorrect)) +    
  geom_boxplot()                  # summarise data with boxplots
```

2. Label the _x_-axis as "Learning condition" and the _y_-axis as "Proportion correct answers". I don't provide the commands for doing so; you already know them from last week.

## Adjusting the appearance of plots

1. By default, `ggplot2` (the part of the `tidyverse` that takes care of plotting) produces plots with a grey background
(which doesn't seem to please anyone save for its developer).
This can be changed by appending the command `theme_bw()` (_black and white_)
to the `ggplot` call:

```{r, fig.asp = 0.5}
ggplot(data = d_box,           
       aes(x = LearningCondition, 
           y = PropCorrect)) +    
  geom_boxplot() +               
  xlab("Learning condition") +          
  ylab("Proportion correct answers") + 
  theme_bw()
```

For other options, see https://ggplot2.tidyverse.org/reference/ggtheme.html.

2. To change the font size, insert a number between the brackets of the `theme_bw()` command. Finding a suitable font size is a matter of trial and error.

```{r, fig.asp = 0.5}
ggplot(data = d_box,           
       aes(x = LearningCondition, 
           y = PropCorrect)) +    
  geom_boxplot() +               
  xlab("Learning condition") +          
  ylab("Proportion correct answers") + 
  theme_bw(15)
```

3. When you're happy with a plot, you can save it using `ggsave()`. You can set the width and height of the figure using the `width` and `height` parameters, e.g.,

```{r, eval = FALSE}
ggsave(here("figures", "03_boxplot.png"), width = 8, height = 4)
```

Per default, the width and height numbers are in inches. See the help page (https://ggplot2.tidyverse.org/reference/ggsave.html) on how to override this.

## Computing summaries
While this is a tutorial on drawing graphs, it's useful at this stage
to know how to obtain some basic information about the datasets you're 
working with. There are many ways to accomplish most things in `R`,
and computing summaries is no exception. The alternative shown here is similar
to the way plots are constructed in `ggplot2`, so hopefully it's easy to
pick up.

First, let's compute the number of observations in the `d_box` dataset
as well as the mean of the `PropCorrect` variable. 
The _pipe_ (`|>`) takes the object to its left and feeds it to the function
to its right (or below). The code below can
be read as follows: 

* _Take `d_box`, THEN_
* _summarise it as follows:_
  - _compute the number of observations (`n()`) and store it as `number`_
  - _and compute the mean of `PropCorrect` and store it as `mean_PropCorrect`._

```{r}
d_box |> 
  summarise(number = n(),
            mean_PropCorrect = mean(PropCorrect))
```

Thus we see that there are 80 observations in the dataset with a mean PropCorrect
of about 0.37.

If we want to compute the mean for the two learning conditions separately,
we insert a corresponding `group_by()` statement. The code below can thus be read
as

* _Take `d_box', THEN_
* _Group it by the values of `LearningCondition`, THEN_
* _For each group created, create a summary containing_
  - _the mean of the `PropCorrect` variable._

```{r}
d_box |> 
  group_by(LearningCondition) |> 
  summarise(mean_PropCorrect = mean(PropCorrect))
```

The word _THEN_ is represented by a so-called pipe operator (|>),
a shortcut for which is ctrl+shift+m on Windows/Linux and cmd + shift + m on Mac.

## Exercise

The file `Vanhove2014_CognateProfile.csv` contains data from @Vanhove2014.
163 Swiss-Germans attempted to translate 45 written Swedish words, all of which
had German, English and/or French cognates.
To check which participants already had some minimal knowledge of Swedish,
they also had to translate five words without cognates in German, English or French
(e.g., _älska_ 'to love'); these are known as 'profile words'.
For the subsequent analysis, it was important to know
(a) how many participants could translate at least one such profile word correctly; and 
(b) if participants who could translate at least one such profile word did better at translating cognates than those who couldn't.

The column `CorrectCognates` contains the number of correct cognate
translations per participant; the column shows whether the participant
in question was able to translate at least one profile word correctly (`yes` vs. `no`).

1. In a new script, read in the dataset into R.
2. Compute the number of participants who were able to correctly translate at least one profile word.
2. Use a boxplot to compare the number of correctly translated cognate words between participants that could translate at least one profile word correctly and those that couldn't. Label your axes appropriately.
3. Compile the HTML report.
4. Submit both the boxplot and the HTML report.
