# Boxplots deluxe

## Goals

1. You'll learn how to add the individual data points underlying a boxplot to this boxplot. This makes for more informative graphs, particularly for smallish datasets (up till 100--150 observations or so).
2. To do this, we'll use a technique known as _jittering_ to make overlapping datapoints more visible.

## Tutorial

Plain boxplots are useful when 
you want to get a quick idea of the distribution
of the data in different groups (see last week).

However, **identically looking boxplots can represent markedly different distributions!** 
See my blog post [_Different distributions, same boxplot_](https://janhove.github.io/posts/2026-08-17-boxplot-distributions/)
for examples.

For this reason, it can be useful to draw boxplots
that also show the _individual_ data points rather
than merely the distribution's quartiles.
Such visualisations are particularly useful
when the dataset isn't prohibitively large,
so that you can actually make out the different data points.

In the following, I assume you'll work within your R project
and that you have the `tidyverse` and `here` packages. Always
use scripts rather than typing your commands directly to the prompt.

1. Read in the dataset `VowelChoices_ij.csv` (see last week). 
2. Draw the boxplot of last week's tutorial again.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
d_box <- read_csv(here("data", "VowelChoices_ij.csv"))
```

```{r, fig.asp = 0.9, fig.width = 4}
ggplot(data = d_box,
       aes(x = LearningCondition, y = PropCorrect)) +
  geom_boxplot() +
  xlab("Learning condition") +
  ylab("Proportion correct answers") +
  theme_bw(12)
```

3. We can add another _layer_ to this graph by
inserting the command `geom_point()` into the ggplot call. 
`geom_point()` draws the data points as, well, points and doesn't summarise its quartiles like `geom_boxplot()` does. 
Since we insert `geom_point()` _after_ `geom_boxplot()`, 
these points will be plotted _on top of_ (rather than underneath) the boxplots.

```{r, fig.asp = 0.9, fig.width = 4}
ggplot(data = d_box,
       aes(x = LearningCondition, y = PropCorrect)) +
  geom_boxplot() +
  geom_point() +    # draw individual data points
  xlab("Learning condition") +
  ylab("Proportion correct answers") +
  theme_bw(12)
```

4. If you count the number of points in this graph, you'll notice that there are fewer points (28) than there were participants (80). 
The reason is that several participants obtained the same result, and their data are plotted on top of each other. 
For this reason, it makes sense to move the plotted points a bit apart (_jittering_). 
Here we can move the points apart horizontally. 
To do this, we need to specify the `position` parameter inside the `geom_point()` command.

    + We can specify both `width` (horizontal jittering) and `height` (vertical jittering). Fiddle with the setting for `width` to see what happens if you change it. I'd leave the value for `height` set to 0 (= no vertical jittering) so that we don't depict proportions of, say, 0.24 as 0.28.
    + Notice that I've additionally specified the `outlier.shape` parameter in the `geom_boxplot()` command as `NA` (_not available_). This prevents the `geom_boxplot()` command from plotting outlying data points so that these points aren't plotted twice (as part of the boxplot and as an individual data point).

```{r, fig.asp = 0.9, fig.width = 4}
ggplot(data = d_box,
       aes(x = LearningCondition, 
           y = PropCorrect)) +
  geom_boxplot(outlier.shape = NA) + # don't plot outliers twice
  geom_point(position = position_jitter(width = 0.2,   # horizontal jittering
                                        height = 0)) + # but no vertical jittering
  xlab("Learning condition") +
  ylab("Proportion correct answers") +
  theme_bw(12)
```

5. To make the individual data points more visually distinct, we can change the plotting symbol by changing the `shape` parameter. This overview shows which plotting symbols are available:

```{r, echo = FALSE, fig.asp = 1/4}
par(bty = "n", cex.lab = 0.6, mar = c(0, 0, 0, 0))
plot(x = 0:25, y = rep(1.1, 26), pch = 0:25, 
     yaxt = "n", ylab = "",
     ylim = c(0.5, 1.5),
     xaxt = "n", xlab = "shape")
text(x = 0:25, y = rep(0.9, 26), labels = as.character(0:25))
```

Empty circles tend to work well (shape 1):

```{r, fig.asp = 0.9, fig.width = 4}
ggplot(data = d_box,
       aes(x = LearningCondition, 
           y = PropCorrect)) +
  geom_boxplot(outlier.shape = NA) + 
  geom_point(position = position_jitter(width = 0.2,   
                                        height = 0), 
             shape = 1) + #  1 = empty circles
  xlab("Learning condition") +
  ylab("Proportion correct answers") +
  theme_bw(12)
```

(Shapes 21 through 25 resemble shapes 1 through 5, but they can be coloured in.)

6. More info: https://janhove.github.io/posts/2016-06-21-drawing-a-boxplot/.

## Exercise

The dataset `Wagenmakers_Zeelenberg.csv` contains data from a study by  @Wagenmakers2016.^[https://osf.io/hgi2y/]
In this experiment, participants were instructed to either `smile` or `pout` (`Condition`)
as they judged the funniness of four cartoons on a 10-point scale (0--9).
The mean rating per participant can be found in the creatively named column `MeanRating`.
The research question was if smiling participants gave higher funniness ratings than pouting ones.

**Your task:** In a new script, compute the number of participants in each condition as well as the mean `MeanRating` per condition. Also draw boxplots (with individual data points) on the basis of which this research question
can be answered. Label your axes appropriately. Save a graph you're happy with
using `ggsave()`. Submit both your graph and the compiled HTML report.
