# Dot plots

## Goals

1. You'll learn how to draw simple dot plots.
2. and why they're better than _pie charts_.
3. You'll gain some experience in debugging R code with typos :)

## Pie charts, bar charts, and dot plots

The three graphs below all show the same data
(the proportion of sales of a pie producer by pie taste).^[Fictitious data from the help page for the R function pie().]

```{r, eval = TRUE, echo = FALSE, fig.height = 4, fig.width = 12}
par(mfrow = c(1, 3), mar = c(8, 4, 2, 2), cex = 1.05)
pie.sales <- c(0.12, 0.3, 0.26, 0.16, 0.04, 0.12)
names(pie.sales) <- c("Blueberry", "Cherry", "Apple", "Boston Cream", "Other", "Vanilla Cream")
pie.sales <- sort(pie.sales)
pie(pie.sales, main = "Pie chart")
par(las = 2)
barplot(pie.sales, main = "Bar chart", ylab = "Proportion of sales")
par(las = 1)
dotchart(pie.sales, xlab = "Proportion of sales", main = "Dot plot", pch = 16)
```

The graph on the left is a traditional **pie chart**.
Though common, this type of graph has serious shortcomings,
for which reason it should rarely be used:

1. Pie charts aren't very flexible: When the number of categories
is larger than here, pie charts are hardly legible.
It's also difficult to add additional information to pie charts
(e.g., the proportion of sales by taste the year before; see below).

2. Readers gauge the numbers (in this case: the proportions)
that the pie chart represents less accurately than when they're 
presented in a bar chart or in a dot plot [@Cleveland1984].

The graph in the middle is a **bar chart**, the one of the right
a **dot plot**. Both are better choices than pie charts.
For one things, they're more flexible.
For instance, you can add the sales from the year before to
both graphs---as separate bars or by using different symbols.
This would be difficult to accomplish in a pie chart:

```{r, eval = TRUE, echo = FALSE, fig.height = 4, fig.width = 10}
par(mfrow = c(1, 2), mar = c(8, 4, 2, 2), cex = 1.05)
pie.sales2015 <- c(0.08, 0.06, 0.18, 0.14, 0.35, 0.19)
names(pie.sales2015) <- names(pie.sales)
par(las = 1)
df.sales <- data.frame(Year2015 = pie.sales2015,
                       Year2016 = pie.sales)
rownames(df.sales) <- names(pie.sales)
par(las = 2)
barplot(height = t(as.matrix(df.sales)), 
        beside = TRUE, legend = TRUE,
        col = c("lightgrey", "grey20"),
        args.legend = list(x = "topleft"),
        ylab = "Proportion of sales", main = "Bar chart")
par(las = 1)
dotchart(pie.sales, xlab = "Proportion of sales", main = "Dot plot", xlim = c(0, 0.35))
points(x = pie.sales2015, y = 1:length(pie.sales2015), pch = 3)
legend("bottomright", pch = c(1, 3), legend = c("2016", "2015"))
par(mfrow = c(1,1))
```

I prefer dot plots because it's easier to visualise a large
number of categories and I find them less 'busy'.

**Tip:** If you prefer bar charts: It's usually better to plot the bars _next_ to each other rather than to stack them on _top_ of each other.

**Tip:** Don't use pie charts.

Incidentally, if you enter the following command in R,
the help page for the `pie()` function opens. 
Then scroll down to the section `Note`.

```{r, eval = FALSE}
# Help page for pie()
?pie
```

## Tutorial

We'll use data from @Vanhove2017 (see Moodle).
For this study, I recruited 175 native speakers of Dutch
from Belgium and the Netherlands. I showed them 44
German words with Dutch cognates 
(e.g., _Stadt_ (NL _stad_) or _Schiff_ (NL _schip_))
and asked them to choose their gender-marked definitive article 
(_der_ (masculine), _die_ (feminine), _das_ (neuter)).
Dutch and German are closely related languages, 
but there are a number of cognates whose grammatical gender
differs between both languages. The Dutch word _strand_ (beach),
for instance, is neuter, whereas the German word _Strand_ is
masculine.
One of the research questions was if Belgian and Dutch participants
were more likely to choose the neuter article _das_ if the German word's
Dutch cognate is neuter than when it has common gender.^[Common gender = non-neuter. Present-day Standard Dutch hardly distinguishes between masculine and feminine gender.]

Read in this dataset in `R`:

```{r, eval = TRUE, echo = TRUE}
dat <- read_csv(here("data", "GermanArticleChoices.csv"))
```

The first few rows should look as follows:

```{r}
head(dat)
```


The column `NeuterResponses` contains the proportion of neuter article
(_das_)
choices per German word (`Stimulus`) per `Country`.
The column `GermanGender` contains the word's correct gender (in German);
the column `DutchGender` contains the grammatical gender of the word's Dutch cognate.

### A first attempt

We can plot the proportion of _das_ choices per word 
separately per country
(Belgium vs. the Netherlands).
This will show us how strongly the response pattern varies per word
and whether Belgian and Dutch speakers of Dutch show different preferences.
The code and comments show how to build this dot plot:

**Tip:** You'll learn more by typing in these commands yourself
and adding your _own_ comments than by copy-pasting them. For
this reason, I added a couple of typos and left out a couple of 
commas and brackets -- all for you to find and fix, enjoy!

```{r, eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE, fig.asp = 1.5}
ggplot(dat,                   
       aes(x = NeuteResponses, 
           y = Stimulus,       
           shape = Country)) + 
  geom_point()                 
```

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.asp = 1.5}
ggplot(dat,
       aes(x = NeuterResponses,
           y = Stimulus,
           shape = Country)) + 
  geom_point() 
```

### Highlight differences between different categories of stimuli

The plot above doesn't offer an answer to the research question
since the difference between stimuli with different Dutch genders
isn't highlighted.
To accomplish this, we can plot stimuli with neuter cognates and
those with common-gender cognates in different boxes (_facetting_,
see the tutorial on line charts).
(To see what `scales = "free_y"` and `space = "free_y"` accomplish, 
leave these parameters out of the call, i.e., using just ` facet_grid(rows = vars(DutchGender))`.)

```{r, eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE, fig.asp = 1.5}
ggplot(dat,
       aes(x = NeuterResponses,
           y = Stimulus 
           shape = Country)) + 
  geom_point() + 
  facet_grid(rows = vars(DutchGender),   # split up vertically by Dutch gender
             scales = "free_y",
             space = "free_y")
```

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.asp = 1.5}
ggplot(dat,
       aes(x = NeuterResponses,
           y = Stimulus, 
           shape = Country)) + 
  geom_point() + 
  facet_grid(DutchGender ~ .,   # vertikal aufteilen je nach ndl. Genus
             scales = "free_y",
             space = "free_y")
```

This graph shows pretty clearly that both Belgian and Dutch
speakers of Dutch pick _das_ more often if the German word
has a neuter-gender cognate than when it has a common-gender
cognate: The points in the lower box lie more to the right
than those in the upper box. With a single exception (_Boot_),
there is no overlap between these two distributions.

Additionally, the responses of Belgian and Dutch participants
don't seem to vary much from one another.
(For instance, we don't observe that most triangles lie to the
right of the circles or that Belgian participants prefer _das_ for _Wurst_
and Dutch participants don't.)

### A final attempt

The previous graph is good enough. But we can do better still.
German is taught in school in both Flanders and the Netherlands,
and at least some participants will have known which words are
neuter in German and which aren't.
So some of the variation between the stimuli will be attributable
to the stimuli's German gender.
To highlight this, we can split up the graph not only by
the words' Dutch gender, but also by their German gender.
And we still have to label the axes!

```{r, eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE,fig.asp = 1.5}
ggplot(dat,
       aes(x = NeuterResponses, 
           y = Stimulus,
           shape = Country)) +
  geom_point() 
  facet_grid(rows = vars(DutchGender, GermanGender), # split up by Dutch and German gender
             scales = "free_y", space = "free_y") +
  xlab("Proportion 'das'") + # axis label for x
  ylab("")                   # the y axis is self-explanatory
```

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE,fig.asp = 1.5}
ggplot(dat,
       aes(x = NeuterResponses, 
           y = Stimulus,
           shape = Country)) +
  geom_point() +
  facet_grid(rows = vars(DutchGender, GermanGender), # aufteilen nach ndl. UND dt. Genus
             scales = "free_y", space = "free_y") +
  xlab("Proportion 'das'") + # Beschriftung x
  ylab("") # y-Achse ist m.E. selbsterklärend
```

The upper three boxes show feminine, masculine and neuter
German words with common-gender cognates; the lower three boxes
show feminine, masculine and neuter German words with neuter-gender
cognates.
The graph shows that the factor _Dutch gender_ is the most important
determinant of the participants' article choices. But the factor
_German gender_ also plays a role: When a German word is neuter,
both Belgian and Dutch people choose _das_ more often than when
it's feminine or masculine.

### Possible improvements and further information

* Now the words are sorted alphabetically in each box. But this order can be customised.
* We could reorder the facets so that the word categories that result in the most _das_ responses are at the top of the graph. While we're at it, we could also give these facets clearer labels.
* The symbols used in the graph above are difficult to distinguish optically. They, too, can be changed.

```{r, eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE,fig.asp = 1.5}
dat <- dat |> 
  # Clearer labels
  mutate(
    word_label = paste0(Stimulus, " (", DutchCognate, ")"),
    dutch_label = paste0("Du.: ", DutchGender),
    german_label = paste0("Gm.: ", GermanGender)
  ) |> 
  # Reorder the labels by proportion 'das'
  # (Run the code with decreasing = FALSE so see what changes)
  mutate(
    word_label = reorder(word_label, NeuterResponses),
    dutch_label = reorder(dutch_label, NeuterResponses, decreasing = TRUE),
    german_label = reorder(german_label, NeuterResponses, decreasing = TRUE)
  )

ggplot(dat,
       aes(x = NeuterResponses,
           y = word_label,
           shape = Country)) +
  geom_point() +
  xlab("Proportion of neuter (das) choices") +
  ylab("German noun (Dutch cognate)") +
  scale_shape_manual(values = c(1, 3)) +
  facet_grid(rows = vars(dutch_label, german_label),
             scales = "free_y",
             space = "free_y") +
  theme_bw()
```


The versatility and usefulness of dotplots in linguistics
is highlighted by @Soenning2016. Another useful introduction
is @Jacoby2006.

I don't expect you to carry out these additional steps. 
But perhaps they're useful to you in the future.

## For other kinds of data
In the tutorial above, we worked with proportions.
But you can use dot plots for other types of data, too.
See http://perceptualedge.com/articles/b-eye/dot_plots.pdf for some examples.

## Exercise

@Baten2011 was interested in the acquisition of the German case system
by Flemish pupils. He reported his results in elaborate tables;
we'll focus on the accuracy data for accusative and dative pronouns
(e.g., _dich_ vs. _dir_) that occur after prepositions.
I tabulated the results in `ErgebniseBaten2011.csv` (Moodle).
This data set contains 4 columns:

* `Grade`: Baten's participants were pupils from 10th, 11th and 12th grade.
* `TargetCase`: The correct form was either an accusative or a dative pronoun.
* `Position`: Some pronouns occurred after a preposition that is either always followed by an accusative (e.g., _bis_) or always followed by a dative (e.g., _nach_); these are marked as `fixed-preposition`. Other pronouns occurred after a preposition that could be followed by both an accusative or a dative, depending on its meaning (e.g., _in_ and _auf_; `wechsel-preposition`).

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 4, fig.height = 4}
baten2 <- read_csv(here("data", "ErgebnisseBaten2011.csv"))
```

Below you'll find two graphs. Both highlight that there's a huge leap in
accuracy between 10th and 11th grade. Moreover, accuracy is higher 
for the 'fixed' prepositions than for the 'Wechselpräpositionen'.
(This is slightly clearer in the second graph, I think.)

```{r, eval = TRUE, echo = FALSE, fig.width = 3, fig.height = 3.1}
ggplot(baten2,
       aes(x = Accuracy,
           y = Grade)) +
  geom_point() +
  facet_grid(Position ~ TargetCase)
```

```{r, eval = TRUE, echo = FALSE, fig.width = 4.5, fig.height = 1.8}
ggplot(baten2,
       aes(x = Accuracy,
           y = Grade,
           shape = Position)) +
  geom_point() +
  facet_grid(. ~ TargetCase)
```

**Your task:** Draw _both_ graphs yourself. Hand in the compiled HTML report.
