# Scatterplots

## Goals

1. You'll learn how to visualise the relationship between two continuous variables in a scatterplot.
2. You'll gain some experience with summarising and manipulating large datasets.

## Summarising datasets

@Vanhove2014 investigated how people's ability to
recognise written and spoken cognates in a related but
unknown language develops throughout the lifespan and
how it is related to linguistic and cognitive factors.
The dataset `Vanhove2014_Translations.csv` contains the
_raw data_ of this study. For each of the 163 participants,
this dataset contains 100 entries (one for each cognate),
for a total of 16,300 rows.
Each translation was rated as correct or incorrect.
Additionally, the dataset contains some information
about the participants (e.g., their performance on other
tasks) as well as about the stimuli (e.g., a measure
expressing its formal similarity to its French, German or English cognate).

```{r}
# Read in data
d <- read_csv(here("data", "Vanhove2014_Translations.csv"))
```

The variables:

* `Stimulus`: The word to be translated.
* `Subject`: The participant's ID.
* `Mode`: Whether the word was presented in its spoken (`Auditory`) or in its written form (`Visual`).
* `Trial`: The position of the word in the task.
* `Translation`: The participant's translation attempt for the stimulus.
* `Correct`: Whether the translation was correct (1) or incorrect (0).
* `Sex`
* `Age`
* `NrLang`: The number of languages the participant spoken.
* `DS.Total`: The participant's score on a working memory task.
* `WST.Right`: The participant's score on a German vocabulary test.
* `Raven.Right`: The participant's score on an intelligence test.
* `English.Total`: The participant's score on an English-language test.
* `Status`: Whether the stimulus has a German, English, or French cognates (`target`) or not (`profile`).
* `MinLev`: The degree of formal discrepancy between the stimulus and its most similar German, English or French cognate. (lower = more similar)

Missing values were labelled `NA` (not available).

A rough summary can be obtained like so:

```{r, eval = TRUE}
summary(d)
```

In order to be able to sensibly visualise these data,
we need to first transform this dataset. If we're interested
in the relationship between the participants' age, sex,
and linguistic and cognitive test results on the one hand
and their translation performance on the other hand,
it seems useful to first compute the number of correct
translations for spoken and written words per participant.
There are a couple of ways you could do this in R;
I find the method outlined below the most transparent one.

Let's start with a simple example. We'll first
compute the total number of correct responses in
the entire dataset. The comments show you how you
can read the R code out loud.

```{r}
# 1st attempt: Total number of correct answers

# Take object 'd' (our dataset), then
d |> 
  # summarise it as follows:
  # - compute the sum of 'Correct' and call this sum 'number_correct'
  summarise(number_correct = sum(Correct))
```

The result is shown in the console. We can also
save it as a separate object so that we can
work with it later. Here I save it as an object
called `overall_accuracy`:

```{r}
overall_accuracy <- d |> 
  summarise(number_correct = sum(Correct))
```

(To type the symbol combination _|>_ (known as the _pipe_; read out loud as 'and then'),
you can also use the shortcut ctrl + shift + m.)

The total number of correct answers in the entire dataset
isn't too useful to us: we need the number of correct answers
per participant.

```{r}
# 2nd attempt: Number of correct answers per participant

# Create new object 'per_part' as follows:
# take 'd', and then...
per_part <- d |> 
  # group it by Subject, and then...
  group_by(Subject)  |>  
  # compute the sum of Correct within each group
  summarise(number_correct = sum(Correct))

# Show result
per_part
```

Now we have the total number of correct translations per participant,
but this figure contains both the responses to written and to spoken words.
It'd be more useful to split this figure up by modality:

```{r}
# 3rd attempt: Number of correct translations per participant,
# split up by modality.

per_part2 <- d |> 
  # group by Subject and Mode
  group_by(Subject, Mode) |> 
  summarise(number_correct = sum(Correct))
per_part2
```

The message (`summarise()` has grouped output by 'Subject'.) isn't harmful, but you can get rid of it as follows:

```{r}
per_part2 <- d |> 
  group_by(Subject, Mode) |> 
  summarise(number_correct = sum(Correct),
            .groups = "drop")
```


Our goal is to visualise the relationship between variables such as
IQ, age, and L1 vocabulary skills on the one hand and the participants'
translation performance on the other hand. To do this, we need to add
these variable to the summary. Again, there are several ways to do this;
here's one.

```{r}
# 4th attempt: Like above, but adding Age, etc.

# Create a new object with participant-related information.
# Take 'd', and then...
part_info <- d |> 
  # Select the following columns
  select(Subject, Age, Sex, WST.Right,
         Raven.Right, English.Total,
         NrLang, DS.Total)
```

Like `d`, `part_info` consists of 16,300 rows.
But each row is repeated 100 times.
Using `distinct()`, the duplicates are deleted:

```{r}
part_info <- part_info |> 
  distinct()
```

Now we can add the information from `part_info` to `per_part2`:

```{r}
# Create new object 'per_part3'.
# Take object 'per_part2', and then...
per_part3 <- per_part2 |> 
  # join it with 'part_info' using 'Subject' as the "zipper"
  full_join(part_info, by = "Subject")

# Show result
per_part3
```

One final problem remains: The number of correct answers
includes responses to both cognates and to non-cognates.
We're only interested in the responses to cognates, however.
Using `filter()`, we only retain the responses that are of interest:

```{r}
# 5th attempt: Like above, but only consider target stimuli.
per_part <- d |> 
  # only consider target stimuli
  filter(Status == "target") |> 
  # group by Subject and Mode
  group_by(Subject, Mode) |> 
  # Compute sum of Correct
  summarise(number_correct = sum(Correct),
            .groups = "drop") |> 
  # Add the information from part_info using Subject as the "zipper"
  full_join(part_info, by = "Subject")
per_part
```

## Drawing scatterplots

To investigate the relationhip between, say, `WST.Right` and `number_correct`,
we can plot these data in a scatterplot. While we're at it, we can split up
this graph into two panels: one for written words, and one for spoken words.

```{r, fig.width = 4.5, fig.height = 3, echo = TRUE}
ggplot(dat = per_part,
       aes(x = WST.Right, 
           y = number_correct)) +
  # Incidentally, you can specify the shape and colour of the points yourself.
  geom_point(shape = 21, colour = "darkblue", fill = "green") +
  # Type in "colours()" to see which colour names are recognised by R.
  # Type in ?pch to see which 'shape' values are accepted.
  xlab("L1 vocabulary test") +
  ylab("Number of correct translations\n(out of 45)") +
  facet_grid(cols = vars(Mode)) +
  labs(title = "A title of your choice",
       subtitle = "(and perhaps even a subtitle)",
       caption = "(and a caption)")
```

(The warning concerns missing values in the `WST` variable. Inspect the dataset
to find for how many participants the `WST` score is missing!)

As we'll discuss in class, scatterplots are the _indispensable_ 
first (and sometimes final) step when analysing 
the relationship between two continuous variables.

As for the decision which variable to put along which axis.
By and large, put the variable that is most likely to be the _cause_
of the relationship along the _x_ axis and the variable that is most
likely to be the _effect_ along the _y_ axis. In this case,
it seems more likely that L1 skills affect one's ability to recognise
cognates in a foreign language than vice versa. Hence, put 
the variable representing L1 skills along the _x_ axis.

## Exercise

Draw a scatterplot that shows the relationship between
the participants' score on the English-language test and the
number of correct translations.
How many participants' data are shown?

Hand in the compiled HTML report.
Mention the number of participants whose data are shown in the R code.
