# Scatterplots with trend lines

## Goal

You'll learn to highlight the trend in the relationship between
two variables using scatterplot smoothers.

## Drawing scatterplot smoothers

```{r, fig.width = 5, fig.height = 2.3, echo = TRUE}
ggplot(dat = per_part,
       aes(x = English.Total, 
           y = number_correct)) +
  geom_point(shape = 1) +
  geom_smooth() + # add smoother
  xlab("Result English test") +
  ylab("Numer of correct translations") +
  facet_grid(cols = vars(Mode))
```

The blue line is the smoother;
the grey ribbon around it is known as a (95%) confidence band
and communicates our uncertainty about the precise location
of the smoother (due to lack of data and variability in the data).
Of course, you can change the colour of the line to suit
your personal taste, and you can turn off the grey ribbon (`se = FALSE`; `se` is short for 'standard error', which is unfortunate -- standard errors and confidence bands aren't the same thing).
We're going to turn off the confidence band because it's bound to be misinterpreted
anyway.

```{r, fig.width = 5, fig.height = 2.3, echo = TRUE}
ggplot(dat = per_part,
       aes(x = English.Total, 
           y = number_correct)) +
  geom_point(shape = 1) +
  geom_smooth(se = FALSE, colour = "red") +
  xlab("Result English test") +
  ylab("Numer of correct translations") +
  facet_grid(cols = vars(Mode))
```

## So what's a scatterplot smoother anyway?

Scatterplot smoothers were developed to discover relationships
(including nonlinear ones) between two variables that aren't
necessarily immediately obvious if the data are shown in a 
scatterplot.
The points on the smoother are a kind of **mean value of the $Y$ variable for the respective $X$ value**.
In the left panel, for instance, the average number of correct translations in the auditory mode for someone with an English test score of 30 is roughly 17--18,
whereas the average number of correct translations for written words for participants with a score of 40 on the English test is about 25.

We needn't amuse ourselves with the maths behind these smoothers,
but the following points are important:

1. The trend line is nearly always a bit _wiggly_ (to use the actual technical term). This is the case even when the relationship itself is as good as linear.

2. The uncertainty about the smoother is larger at the extremes (high and low $x$ values). The confidence band is correspondingly wider at these places. But again, we're not going to plot the confidence bands in order not to confuse ourselves or our readership.

3. The default settings for `geom_smooth()` tend to work fairly well, but sometimes it's necessary to fiddle with them so that the smoother captures the trend in the data better. See next section.

## Examples of nonlinear relationships

This section serves to show you what true nonlinear relationships
can look like and how you sometimes need to fiddle with `geom_smooth()`'s settings.

In the first example, the _red_ line was drawn with the default settings. This line doesn't capture an important feature of the relationship (the data points go up and down). The _blue_ line captures this trend much better. It was drawn using the command `geom_smooth(span = 0.1)`. The `span` parameter determines how wiggly the curve may be (the smaller `span`, the wigglier the curve). By default, span is set to 0.75. Finding a decent span value is matter of trial and error.

```{r, fig.width = 3.5, fig.height = 2.3, echo = FALSE, message = FALSE, warning = FALSE}
df <- data.frame(x = runif(100, -4*pi, 4*pi))
df$y <- sin(df$x) + 0.2*df$x + rnorm(50, sd = 0.1)
ggplot(dat = df,
       aes(x = x, y = y)) +
  geom_point(pch = 1) +
  geom_smooth(colour = "red", se = FALSE, lwd = 0.5,
              method = "loess") +
  geom_smooth(span = 0.1, colour = "blue", se = FALSE, lwd = 0.5,
              method = "loess")
```

In the second example, the _red_ line was drawn
using `geom_smooth(span = 0.1)`.
This line is much too wiggly, and it essentially 
models random deviations from the general trend.
The _blue_ line, drawn with the default setting (`span = 0.75`),
captures the general trend much more sensibly.
The _green_ line, by contrast, isn't wiggly enough (`span = 3`).

```{r, fig.width = 6, fig.height = 2, echo = FALSE,message = FALSE, warning = FALSE}
set.seed(21-11-2016)
df <- data.frame(x = runif(100, -4*pi, 4*pi))
df$y <- 3*df$x^3 - 0.2*df$x^2 - 50*df$x + rnorm(50, sd = 1000)
p1 <- ggplot(dat = df,
       aes(x = x, y = y)) +
  geom_point(pch = 1) +
  geom_smooth(span = 0.1, colour = "red", se = FALSE, lwd = 0.5, method = "loess")
p2 <- ggplot(dat = df,
       aes(x = x, y = y)) +
  geom_point(pch = 1) +
  geom_smooth(span = 0.75, colour = "blue", se = FALSE, lwd = 0.5, method = "loess")
p3 <- ggplot(dat = df,
       aes(x = x, y = y)) +
  geom_point(pch = 1) +
  geom_smooth(span = 3, colour = "darkgreen", se = FALSE, lwd = 0.5, method = "loess")
gridExtra::grid.arrange(p1,p2,p3,ncol=3)
```

**Summing up:** Generally, the default settings work reasonably well. But when you notice that visually salient patterns
in the scatterplot aren't captured by the trend line,
you need to fiddle a bit with the `span` parameter.

More generally, data analysis and statistics aren't a matter of
blindly applying formulae and recipes.

## Exercise

Draw a scatterplot that shows the relationship between the participants' age and their performance on the cognate translation task. How would you describe this relationship?

Now add a smoother (default settings) to this scatterplot.
Does it capture the trend that you identified in the scatterplot?

Now fiddle with the `span` parameter (0.1, 0.2, 0.3 etc.)

Hand in the graph you think works best as well as the compiled
HTML report.
