# Further resources {-}

If you want to learn more about drawing graphs, check out @Healy2019 and @Wilke2019.
Both books are freely available online:

* https://socviz.co/
* https://clauswilke.com/dataviz/

You may also find the tutorials on https://github.com/janhove/DatasetsAndGraphs helpful.

# Common error sources in R {-}

## `object 'x' not found`

You're trying to call an object
that hasn't been loaded yet. This error 
is commonly caused by typos.

Example:

```{r, error = TRUE}
x <- c(1, 4, 5)
mean(X)
```

You've created an object named `x`, but on the second line, you're trying
to call an object named `X`  (capitalisation!).

When in doubt, you can inspect the names of all objects currently loaded
by running this command:

```{r, eval = FALSE}
ls()
```

## `could not find function 'x'`

You're trying to run a function that doesn't exist or that requires you
to load an extension package first. Check whether you've typed in its name
correctly. If it's a function from an extension package, check if that package
has been loaded.


```{r, echo = FALSE}
detach("package:tidyverse", unload=TRUE)
detach("package:ggplot2", unload=TRUE)
```

Example:

```{r, error = TRUE}
ggplot(dat,
       aes(x = predictor,
           y = outcome)) +
  geom_point()
```

You need to load `tidyverse` (or `ggplot2`) before you can use the
`ggplot()` function.

## `there is no package called 'x'`

You're trying to load an extension package that hasn't been installed yet.

Example:

```{r, error = TRUE}
library(ggjoy)
```

Install the package first:

```{r, eval = FALSE}
install.package("ggjoy")
```

## `unexpected symbol`

Check if you've forgotton a comma or bracket somewhere, or if you've used
a comma or bracket that shouldn't be there.

Example:

```{r, message = FALSE, error = TRUE}
library(ggplot2)
ggplot(data = iris
       aes(x = Sepal.Width,
           y = Sepal.Length)) +
  geom_point()
```

This error is difficult to spot because the error message isn't too helpful:
There is a missing comma at the end of line 2, after `iris`.

Example:

```{r, error = TRUE}
ggplot(data = iris)
       aes(x = Sepal.Width,
           y = Sepal.Length)) +
  geom_point()
```

The bracket after `iris` should be a comma.

```{r, error = TRUE}
ggplot(data = iris,
       aes(x = Sepal.Width,
           y = Sepal.Length))) +
  geom_point()
```

Superfluous bracket at the end of line 3.

## It doesn't work and there isn't even an error message

You've probably missed a bracket.

Example: The following command won't produce an error message, but no graphic either:

<pre>
> ggplot(data = iris,
+        aes(x = Sepal.Width,
+            y = Sepal.Length) +
+   geom_point()
+
</pre>

In the console, you'll see that there's a new line starting with `+`.
This means that the previous command hasn't been completed yet:
completed commands are followed by new lines starting with `>`.
In this particular case, the malfunction is caused by a missing
bracket on the third line. The last bracket on this line closes
the command `aes(`, but you need a second bracket to close
the command `ggplot(`.

Here is another example of code that doesn't produce an error
but that doesn't produce the desired graphic either:

```{r, echo = FALSE}
library(ggplot2)
```


```{r}
ggplot(data = iris,
       aes(x = Sepal.Width,
           y = Sepal.Length))
geom_point()
```

Reason: the `+` on line 3 is missing.

## `object of type 'closure' is not subsettable`
Ah, the quintessential R error message. It can be generated 
in a number of ways, but you're probably accidentally treating
a _function_ as a _variable_. Here's an example: Let's say you
want to compute the mean of the numbers 200 and 800. One
way to do so is like so:

```{r}
mean(c(200, 800))
```

However, if you've accidentally used square brackets, the error
pops up:

```{r, error = TRUE}
mean[c(200, 800)]
```

What you're actually doing with the square brackets is reading
out two values of `mean`, namely the 200th and 800th value it's holding.
However, `mean` doesn't hold any values; it's a function, not a variable.

To fix the problem, go through your code and check if you haven't mixed up
square and round brackets.



# Software versions {-}

```{r, echo = FALSE}
devtools::session_info()
```

`r if (knitr::is_html_output()) '
# References {-}
'`
